---
title: "ENGSCI 255 Lab 1"
author: "Navindd Raj"
date: "Due Date: 5pm Monday 7 May 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library("ggplot2")
library("rpart")
library("rpart.plot")
library("rattle")
library("gridExtra")
theme_update(plot.title = element_text(hjust = 0.5))
theme_update(plot.subtitle = element_text(hjust = 0.5))
```

##Question 1

####Read in and inspect the data
```{r, fig.height=3.75}
titanic.df <- read.csv("titanic.csv",header=TRUE)

# make "Survived" and "Pclass" into factors, as they are currently numerical
titanic.df = within(titanic.df, {SurvivedFactor = factor(Survived)})
titanic.df = within(titanic.df, {PclassFactor = factor(Pclass)})

```

##Question 3

```{r}
set.seed(50) # sets a seed for random number generator
```

####a)

Randomly select 125 survivors and 125 nonsurvivors for our training set:
```{r}
# order the survivors (0 then 1)
ordered_titanic = titanic.df[order(titanic.df$Survived),] # our data set to be worked on
# we know there are 342 survivors, so 891-342=549 nonsurvivors
train = c(sample(1:549,125),sample(550:891,125)) # training subset of 250 passengers
```

Now, we can use rpart to create a tree, using only "Sex" as an independent attribute. Our dependent attribute is "Survived", as we want to predict this attribute for other data sets.

```{r, fig.align='center'}
tree=rpart(SurvivedFactor~Sex,data=ordered_titanic,subset=train) # create tree
fancyRpartPlot(tree, sub = "") # display tree
```

In the tree above, the uppermost number represents the dominant value for "Survived" attribute at that node, so 0 if there are more nonsurvivors than survivors at the current node and 1 for vice versa. The first node doesn't matter since it's a 50/50 split.

The middle numbers represent the proportion of dead/survived datapoints in a particular node (assuming the dependent attribute in the tree is "Survived")

The lowest number represents the percentage of datapoints at the current node vs the entire dataset the tree was built on.

####b)

Now we want to predict the "Survived" attribute of passengers NOT in the training subset (so use the dataframe minus the training subset for our test set).

```{r}
# display prediction information in a table
# ordered_titanic[-train] returns rows in the ordered titanic dataset that is not indexed in the train vector, i.e. our test set
table(predict(tree,ordered_titanic[-train,],type="class"),ordered_titanic[-train,"Survived"])
```

The predictions for the test set based on this particular tree is shown in a table form above.

The rows represent if the "Survived" attribute is 0 or 1, i.e. if the passenger in the test set were predicted to survived or not. The columns represent if that particular passenger actually survived (1) or not (0).

For this tree formed by only splitting males and females, $360/(360+68) = 84.1\%$ of nonsurvivors were predicted correctly, and $149/(149+64) = 70.0\%$ of survivors were predicted correctly, which is a decent performance for a tree with only one split.

####c) Create four additional classification tress:

#####i. With "Pclass", "Age", "Sibsp" and "Parch" and "Sex"

We use the same training set as before, so we can create our tree using rpart straight away.

```{r, fig.align='center'}
tree2=rpart(SurvivedFactor~Pclass+Age+SibSp+Parch,data=ordered_titanic,subset=train) # create tree
fancyRpartPlot(tree2, sub = "") # display tree
```

```{r}
# display prediction information in a table
# ordered_titanic[-train] returns rows in the ordered titanic dataset that is not indexed in the train vector, i.e. our test set
table(predict(tree2,ordered_titanic[-train,],type="class"),ordered_titanic[-train,"Survived"])
```

with sex
```{r, fig.align='center'}
tree1=rpart(SurvivedFactor~Pclass+Age+SibSp+Parch+Sex,data=ordered_titanic,subset=train) # create tree
fancyRpartPlot(tree1, sub = "") # display tree
```

```{r}
# display prediction information in a table
# ordered_titanic[-train] returns rows in the ordered titanic dataset that is not indexed in the train vector, i.e. our test set
table(predict(tree1,ordered_titanic[-train,],type="class"),ordered_titanic[-train,"Survived"])
```


